---
title: "Lab 09"
author: "Liz Chamiec-Case"
format: pdf
editor: visual
---

## 

PDB Statistics

**Q1:** What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.

X-Ray: 169,794

EM: 12,835

X-Ray and EM: 182,629

**Q2:** What proportion of structures in the PDB are protein?

171,221 / 196,779

**Q3:** Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

1988

**Q4**: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

Two of the atoms in a water molecule are hydrogens. For simplicity, this structure does not show hydrogen molecules to reduce clutter in the visualization.

**Q5**: There is a conserved water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have (see note below)?

HOH308:O

Q6: As you have hopefully observed HIV protease is a homodimer (i.e. it is composed of two identical chains). With the aid of the graphic display and the sequence viewer extension can you identify secondary structure elements that are likely to only form in the dimer rather than the monomer?

alpha helix, beta sheet

FINISH THIS QUESTION

```{r}
library(bio3d)
pdb <- read.pdb("1hsg.pdb")
pdb
```

Q7: How many amino acid residues are there in this pdb object?

198

Q8: Name one of the two non-protein residues?

HOH, MK1

Q9: How many protein chains are in this structure?

2

```{r}
attributes(pdb)
```

```{r}
head(pdb$atom)
```

```{r}
# Install packages in the R console not your Rmd, commenting out after installation

# install.packages("bio3d")
# install.packages("ggplot2")
# install.packages("ggrepel")
# install.packages("devtools")
# install.packages("BiocManager")
# 
# BiocManager::install("msa")
# devtools::install_bitbucket("Grantlab/bio3d-view")
```

Q10. Which of the packages above is found only on BioConductor and not CRAN?

msa

Q11. Which of the above packages is not found on BioConductor or CRAN?

bio3d-vew from Grantlab github

Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

true

```{r}
library(bio3d)
aa <- get.seq("1ake_A")
aa
```

Q13. How many amino acids are in this sequence, i.e. how long is this sequence?

214 amino acids

```{r}
# Blast or hmmer search 
b <- blast.pdb(aa) 
```

```{r}
# Plot a summary of search results
hits <- plot(b)
```

```{r}
# List out some 'top hits'
head(hits$pdb.id)
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
# Vector containing PDB codes for figure axis
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
# plot(pdbs, labels=ids) # won't render with this uncommented
```

```{r}
# annotate collected PDB structures
anno <- pdb.annotate(ids)
unique(anno$source)
```

```{r}
anno
```

```{r}
# perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
# calculate RMSD
rd <- rmsd(pdbs)

# structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd,k=3)

plot(pc.xray,1:2,col="grey50",bg=grps.rd,pch=21,cex=1)
```

```{r}
#Plotting results with ggplot2
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```

```{r}
# NMA of all structures
modes <- nma(pdbs)
plot(modes, pdbs, col=grps.rd)
```

**Q14.** What do you note about this plot? Are the black and colored lines similar or different? Where do you think they differ most and why?

The black and colored lines differ quite a bit. It always occurs when the colored bar at the top/bottom is white instead of light or dark gray. This might be because these are areas where the structures differ.
