---
title: "pertussis mini project"
author: "Liz Chamiec-Case"
format: pdf
editor: visual
---

## Investigating pertussis cases by year

```{r}
library(datapasta) 

cdc <- tibble::tribble(
                               ~Year,    ~Reported_Cases,
                                1922,	107473,
                                    1923,	164191,
                                    1924,	165418,
                                    1925,	152003,
                                    1926,	202210,
                                    1927,	181411,
                                    1928,	161799,
                                    1929,	197371,
                                    1930,	166914,
                                    1931,	172559,
                                    1932,	215343,
                                    1933,	179135,
                                    1934,	265269,
                                    1935,	180518,
                                    1936,	147237,
                                    1937,	214652,
                                    1938,	227319,
                                    1939,	103188,
                                    1940,	183866,
                                    1941,	222202,
                                    1942,	191383,
                                    1943,	191890,
                                    1944,	109873,
                                    1945,	133792,
                                    1946,	109860,
                                    1947,	156517,
                                    1948,	74715,
                                    1949,	69479,
                                    1950,	120718,
                                    1951,	68687,
                                    1952,	45030,
                                    1953,	37129,
                                    1954,	60886,
                                    1955,	62786,
                                    1956,	31732,
                                    1957,	28295,
                                    1958,	32148,
                                    1959,	40005,
                                    1960,	14809,
                                    1961,	11468,
                                    1962,	17749,
                                    1963,	17135,
                                    1964,	13005,
                                    1965,	6799,
                                    1966,	7717,
                                    1967,	9718,
                                    1968,	4810,
                                    1969,	3285,
                                    1970,	4249,
                                    1971,	3036,
                                    1972,	3287,
                                    1973,	1759,
                                    1974,	2402,
                                    1975,	1738,
                                    1976,	1010,
                                    1977,	2177,
                                    1978,	2063,
                                    1979,	1623,
                                    1980,	1730,
                                    1981,	1248,
                                    1982,	1895,
                                    1983,	2463,
                                    1984,	2276,
                                    1985,	3589,
                                    1986,	4195,
                                    1987,	2823,
                                    1988,	3450,
                                    1989,	4157,
                                    1990,	4570,
                                    1991,	2719,
                                    1992,	4083,
                                    1993,	6586,
                                    1994,	4617,
                                    1995,	5137,
                                    1996,	7796,
                                    1997,	6564,
                                    1998,	7405,
                                    1999,	7298,
                                    2000,	7867,
                                    2001,	7580,
                                    2002,	9771,
                                    2003,	11647,
                                    2004,	25827,
                                    2005,	25616,
                                    2006,	15632,
                                    2007,	10454,
                                    2008,	13278,
                                    2009,	16858,
                                    2010,	27550,
                                    2011,	18719,
                                    2012,	48277,
                                    2013,	28639,
                                    2014,	32971,
                                    2015,	20762,
                                  2016,	17972,
                                  2017,	18975,
                                  2018,	15609,
                                  2019,	18617
    )

cdc
```

\

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(x=Year, y=Reported_Cases) +
  geom_point() +
  geom_line() +
  labs(x="Year",y="Number of Cases",title="Pertussis Cases by Year") +
  geom_vline(xintercept=1946,color="blue") +
  geom_vline(xintercept=1996,color="red")
```

Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the aP vaccine was introduced, the number of cases started to rise for the first time in years. This could be due to distrust in a new vaccine, development of new strains that the newer vaccines are less effective against, or better testing methods.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc...)?

```{r}
table(subject$race,subject$biological_sex)
```

```{r}
library(lubridate)

# i) average age of wP individuals
wP <- subset(subject,infancy_vac == "wP")
wP['age_days'] <- today() - ymd(wP$year_of_birth)
wP['age_years'] <- round(time_length( today() - ymd(wP$year_of_birth),  "years"))
summary(wP$age_years)
```

```{r}
# ii) average age of aP individuals
aP <- subset(subject,infancy_vac == "aP")
aP['age_days'] <- today() - ymd(aP$year_of_birth)
aP['age_years'] <- round(time_length( today() - ymd(aP$year_of_birth),  "years"))
summary(aP$age_years)
```

iii\) are they significantly different?

In the absence of a statistical test, we can tell that the average ages are likely significantly different because there is no overlap in the data (max(aP) \< min(wP)). This makes sense since the wP vaccine was likely discontinued when the aP vaccine was introduced.

also significant with a wilcox test:

```{r}
x <- t.test(wP$age_years,aP$age_years)

x$p.value
```

Q8. Determine the age of all individuals at time of boost?

```{r}
subject['age_at_boost'] <- round(time_length(ymd(subject$date_of_boost) - ymd(subject$year_of_birth),"years"))
subject['age_at_boost']
```

Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Yes - aP has a much smaller mean and variance whereas wP has a higher mean and variance. There is almost no overlap between the two groups.

```{r}
# Or use wilcox.test() 
x <- t.test(wP$age_years,aP$age_years)

x$p.value
```

## Joining multiple tables

```{r}
# Complete the API URLs
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE)

# join datasets
library(dplyr)
meta <- full_join(specimen, subject)
dim(meta)
head(meta)
```

```{r}
# join meta and titer
abdata <- inner_join(titer, meta)
dim(abdata)
```

```{r}
table(abdata$isotype)
```

Q12. What do you notice about the number of visit 8 specimens compared to other visits?

```{r}
table(abdata$visit)
```

There are substantially fewer visit 8 specimens as compared to the other visits

## Examine IgG1 Ab titer levels

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)
```

Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

FIM2/3 differs quite a bit over time, although it seems to have a lot of outliers. It also saturates the plot, so it's difficult to tell if other antigens are responding similarly. This trend is not well reflected in the CBI-PM data, though.

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a "control" antigen ("Measles", that is not in our vaccines) and a clear antigen of interest ("FIM2/3", extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).

```{r}
filter(ig1, antigen=="TT") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

Q16. What do you notice about these two antigens time courses and the FIM2/3 data in particular?

Both TT and FIM2/3 antigen time courses rise over time, perhaps TT more so than FIM2/3 but difficult to tell because TT has large error bars. FIM2/3 seems to peak around visit 5 whereas TT does not come back down.

Q17. Do you see any clear difference in aP vs. wP responses?

In FIM2/3, aP seems to increase more than wP does over time. The effect is smaller if present at all in TT.

## Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 

ssrna <- inner_join(rna, meta)
```

```{r}
ggplot(ssrna) +
  aes(x=visit, y=tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

Q19.: What do you notice about the expression of this gene (i.e. when is it at it's maximum level)?

The expression of this gene peaks very specifically at visit 4 by more than double the expression at any other visits.

Q20. Does this pattern in time match the trend of antibody titer data? If not, why not?

This makes sense - the antibody titer data for FIM2/3 peaks at visit 5 then declines after that. It takes time from expression of the gene to produce/circulate the protein so, depending on how far apart the visits are, we might expect to see this.

```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```
