---
title: "class10_halloween"
format: pdf
editor: visual
---

```{r}
# import data

candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

Q1. How many different candy types are in this dataset?

```{r}
print(paste('there are',nrow(candy),'candy types in the dataset'))
```

Q2. How many fruity candy types are in the dataset?

```{r}
print(paste('there are',sum(candy$fruity),'fruity candy types in the dataset'))
```

Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
candy["Almond Joy", ]$winpercent
```

Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
# introduce skimr

library("skimr")
skim(candy)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent is out of 100 instead of out of 1

Q7. What do you think a zero and one represent for the candy\$chocolate column?

0 = candy does not have chocolate; 1 = candy does have chocolate (binary code)

Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?

no - there is a right hand tail

Q10. Is the center of the distribution above or below 50%?

below

Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
print(paste('chocolate:',mean(candy$winpercent[as.logical(candy$chocolate)]),'; fruity:',mean(candy$winpercent[as.logical(candy$fruity)])))
print('on average, chocolate candy is higher ranked than fruity candy')
```

Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```

## Overall Candy Rankings

Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent,decreasing=TRUE),], n=5)
```

Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy,aes(winpercent, rownames(candy))) +
  geom_col()
```

Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() 
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

Q17. What is the worst ranked chocolate candy?

Charleston Chew

Q18. What is the best ranked fruity candy?

Starburst

## Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip is least popular

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate; pluribus and bar

Q23. Similarly, what two variables are most positively correlated?

bar and chocolate; chocolate and winpercent

## PCA

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
# plot PC1 and PC2
plot(pca$x[,1:2],col=my_cols,pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
#ggplotly(p)
```

```{r}
# make loadings plot

par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

This makes sense since fruity candies are more likely than chocolate, caramel, or nutty candies to be hard candies and pluribus. Thus, these three features are likely to describe the same candies.
