---
title: "Vaccination Rate Miniproject"
author: "Liz Chamiec-Case"
format: pdf
editor: visual
---

## Getting Started

```{r}
# Import vaccination data
vax <- read.csv('covid19vaccinesbyzipcode_test.csv')
head(vax)
```

Q1. What column details the total number of people fully vaccinated?

"persons_fully_vaccinated"

Q2. What column details the Zip code tabulation area?

"zip_code_tabulation_area"

Q3. What is the earliest date in this dataset?

"2022-11-22"

Q4. What is the latest date in this dataset?

"2021-01-05"

```{r}
# skim dataset

skimr::skim(vax)
```

Q5. How many numeric columns are in this dataset?

13

Q6. Note that there are "missing values" in the dataset. How many NA values there in the persons_fully_vaccinated column?

15440

Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

0.09

Q8. \[Optional\]: Why might this data be missing?

Could be unaccounted for due to sources of vaccine not counted by this data (e.g. military, etc.)

## Working with Dates

```{r}
library(lubridate)
today()
```

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[1]
```

```{r}
# date span of dataset

vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[length(vax$as_of_date)]
```

Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

## Working with Zip Codes

```{r}
library(zipcodeR)
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

```{r}
# Pull data for all ZIP codes in the dataset
# zipdata <- reverse_zipcode( vax$zip_code_tabulation_area ) # takes too long
```

### San Diego Area

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
nrow(sd)
```

Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd[which.max(sd$age12_plus_population),]
```

92154

```{r}
# San Diego county on 2022-11-15

library(dplyr)

sd.nov15 <- filter(vax, county == "San Diego", as_of_date == '2022-11-15')
head(sd.nov15)
```

Q13. What is the overall average "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2022-11-15"?

```{r}
mean(sd.nov15$percent_of_population_fully_vaccinated, na.rm=TRUE) # na values are removed
```

Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2022-11-15"?

```{r}
library(ggplot2)

ggplot(sd.nov15) + 
  geom_histogram(aes(x = percent_of_population_fully_vaccinated))
```

## La Jolla/UCSD

```{r}
# isolate La Jolla data and print population age 5+ on first day of data
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

```{r}
# time rate course for vaccination at UCSD

ggplot(ucsd) +
  aes(x = as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-11-15")

head(vax.36)
```

Q16. Calculate the mean "Percent of Population Fully Vaccinated" for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2022-11-15". Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
avg.vaccinated <- mean(vax.36$percent_of_population_fully_vaccinated)

ggplot(ucsd) +
  aes(x = as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept=avg.vaccinated,linetype=2,color="red") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the "Percent of Population Fully Vaccinated" values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date "2022-11-15"?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  geom_histogram(aes(x=percent_of_population_fully_vaccinated))
```

Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

92040 less than calculated above

```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

92109 less than calculated above

Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population \> 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
vax.36.all.mean <- mean(vax.36.all$percent_of_population_fully_vaccinated, na.rm=TRUE)


ggplot(vax.36.all) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x='Date', y='Percent Vaccinated',
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = vax.36.all.mean, linetype=2)
```

Q21. How do you feel about coming to class in person after Thanksgiving break?

Honestly probably won't come, but that's more because I prefer to do the labs on my own. I appreciate your asking though!
